---
title: "Getting started with `SoundShape` package"
author: "Pedro Rocha"
date: "October 2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Welcome to `SoundShape`

Here, you will find information on how to implement a promising, and yet little explored method for biacoustical analysis: the so called *eigensound* analysis developed by [MacLeod, Krieger and Jones (2013)](https://doi.org/10.4404/hystrix-24.1-6299).

Eigensound is a multidisciplinary method focused on the direct comparison between homologous sounds from different species (*i.e.* stereotyped calls/acoustic units; Macleod *et al.*, 2013; Rocha & Romano *in prep*). It consists on applying a sampling grid over the representation of sound (*i.e.* spectrogram data; Figs. 1 and 2) and then translate the spectrogram into a dataset that can be analyzed similarly to coordinate sets used in Geometric Morphometrics Methods (GMM). By doing so, eigensound crosses the bridge between Bioacoustics and GMM.

Despite being well described by Macleod *et al.* (2013), the method lacked a free and open platform to run the analysis. `SoundShape` package was written on [R platform](https://www.r-project.org/) to fill this applicability gap. The package features functions that enable anyone familiar with `R` to easily go from sound waves to principal components analysis (PCA), using tools extracted from traditional bioacoustics (*i.e.* [tuneR](https://CRAN.R-project.org/package=tuneR) and [seewave](https://CRAN.R-project.org/package=seewave) packages), geometric morphometrics (*i.e.* [geomorph](https://CRAN.R-project.org/package=geomorph) package) and multivariate analysis (*e.g.* [stats](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/00Index.html) package).

Thanks for using `SoundShape` and enjoy your reading!
  
  **Note:** Should you experience problems running any function, please feel free to [report any issues here](https://github.com/p-rocha/SoundShape/issues). 
  
```{r Fig1-3D-spectros, fig.align='center', fig.height=3.5, fig.width=7.2}
library(SoundShape)

# Sample data from SoundShape
data(cuvieri)

# Select acoustic unit from sample
cuvieri.cut <- seewave::cutw(cuvieri, f=44100, from = 0.05, to=0.45, output="Wave")

# 3D spectrogram
par(mfrow=c(1,2), mar=c(0,2,1,0))
threeDspectro(cuvieri.cut, flim=c(0, 2.5), 
              colkey=list(plot=FALSE), cex.axis=0.4, cex.lab=0.8, resfac=3)

# Semilandmarks from sampled surface
threeDspectro(cuvieri.cut, flim=c(0, 2.5), plot.type="points",
              samp.grid=TRUE, x.length=70, y.length=50, main="Semilandmarks 3D",
              colkey=list(plot=FALSE), cex.axis=0.4, cex.lab=0.8)

```

**Figure 1:** Graphical outputs using `threeDspectro` function from `SoundShape` package: (left) 3D spectrogram and (right) points (*i.e.* semilandmarks) sampled from 3D spectrogram data. `cuvieri` sample from `SoundShape` package. 

```{r Fig2-2D-spectros, fig.align='center', fig.width=7.2, fig.height=3.5}
# Traditional oscillogram and spectrogram
par(mfrow=c(1,2), mar=c(4,4,2,1)) # view side by side
seewave::oscillo(cuvieri.cut, title="Oscillogram")
seewave::spectro(cuvieri.cut, flim=c(0, 2.5), grid=FALSE, scale=FALSE, main="Spectrogram")
```

**Figure 2:** Graphical outputs using `seewave` package: (left) Oscillogram created with `oscillo` function and (right) 2D spectrogram created with `spectro` function. `cuvieri` sample from `SoundShape` package. 

## Installation
`SoundShape` package is available on [R platform](https://www.r-project.org/) as a development version from [GitHub](https://github.com/). In order to download it, make sure to have already installed an updated `R` version (3.3.1 or above) and [devtools](https://CRAN.R-project.org/package=devtools) package.

```r
# install.packages("devtools")
devtools::install_github("p-rocha/SoundShape")
```

## Citation
In case you wish to use and cite `SoundShape` package, use `citation("SoundShape")`.

```{r citation}
citation("SoundShape")
```


# Workflow using `SoundShape` package

## 1. Definition of homologous acoustic units

Since eigensound is centered around stereotyped acoustic units, the foremost step in sound shape study is the careful definition of units from which analysis will be conducted. Although there is no universal concept of a homologous unit of biological sound encompassing the majority of calling organisms, each higher taxon has its own approaches for homologous sound comparison (see Rocha & Romano *in prep* for details on homology between units). Herein, we focus on stereotyped calls from three frog species: *Physalaemus centralis*, *P. cuvieri* and *P. kroyeri* (`centralis`, `cuvieri` and `kroyeri` sample datas, respectively; Figs. 3 -- 5). When dealing with real-life data, undertake literature research before defining the comparable acoustic units.

```{r Fig3-centralis-units, fig.align='center', fig.height=3, fig.width=7.2} 
# Samples of data from SoundShape package
data(cuvieri)
data(centralis)
data(kroyeri)

# Plot spectro from sample and highlight acoustic units

# centralis
seewave::spectro(centralis, flim = c(0, 4), wl=512, f=44100, ovlp=70, grid=FALSE)
graphics::abline(v=c(0.1, 0.8, 1.08, 1.78, 2.1, 2.8), lty=2)
```

**Figure 3:** Spectrogram image of `centralis` sample (`SoundShape` package), containing a sequence of three stereotyped vocalizations, each representing a comparable acoustic unit.  

```{r Fig4-cuvieri-units, fig.align='center', fig.height=3, fig.width=7.2}
# cuvieri
seewave::spectro(cuvieri, flim = c(0,4), wl=512, f=44100, ovlp=70, grid=FALSE)
graphics::abline(v=c(0.05, 0.45, 0.73, 1.13, 1.47, 1.87), lty=2)
```

**Figure 4:** Spectrogram image of `cuvieri` sample (`SoundShape` package), containing a sequence of three stereotyped vocalizations, each representing a comparable acoustic unit. 

```{r Fig5-kroyeri-units, fig.align='center', fig.height=3, fig.width=7.2}
# kroyeri
seewave::spectro(kroyeri, flim = c(0, 4), wl=512, f=44100, ovlp=70, grid=FALSE)
graphics::abline(v=c(0.16, 0.96, 1.55, 2.35, 2.9, 3.8), lty=2)
```

**Figure 5:** Spectrogram image of `kroyeri` sample (`SoundShape` package), containing a sequence of three stereotyped vocalizations, each representing a comparable acoustic unit. 


## 2. Create folder to store acoustic units

`eigensound` function (`SoundShape` package) focus on the acquisition of point coordinates (*i.e.* semilandmarks) from multiple `".wav"` files (WAV: [Waveform Audio File Format](https://en.wikipedia.org/wiki/WAV)), with each file representing a comparable acoustic unit (see section 1). These `".wav"` files must be stored on the same folder somewhere in your computer, which can be created manually at your console and subsequently assigned as [working directory in R](http://rprogramming.net/set-working-directory-in-r/). 

Alternatively, the codes below can be used to create a folder at *e.g.* the current working directory, and a subfolder to store the upcoming outputs from `eigensound`:
  
```r
# Create folder to store ".wav" files
wav.at <- file.path(getwd(), "example SoundShape")
dir.create(wav.at)

# Create subfolder to store results
store.at <- file.path(getwd(), "example SoundShape/output")
dir.create(store.at)
```

## 3. Select and store acoustic units as separate `".wav"` files

Once the stereotyped units have been defined, a reasonable number of units should be selected from the sample and stored as new `".wav"` files  on the folder specified by `wav.at` (see section 2 for folder paths). Each `".wav"` file should represent a single acoustic unit selected from the original sound wave. Besides, since the slightest graphical change may incur in biased results (MacLeod et al., 2013),  selection must account for optimal signal to noise ratio (*i.e.* "clean" recording), and no overlapping frequencies from other individuals, species, or background noise. Editing and filtering of sound waves must be restricted to a bare minimum. 

The selection can be performed on numerous softwares of acoustic analysis outside `R` platform (*e.g.* [Audacity](https://www.audacityteam.org/), [Raven Pro](http://ravensoundsoftware.com/software/raven-pro/)), or using some functions from [seewave](https://CRAN.R-project.org/package=seewave) and [tuneR](https://CRAN.R-project.org/package=tuneR) packages as exemplified below:
  
```{r cut-units}
# Select acoustic units
cut.centralis <- seewave::cutw(centralis, f=44100, from=0, to=0.9, output = "Wave")
cut.cuvieri <- seewave::cutw(cuvieri, f=44100, from=0, to=0.9, output = "Wave")
cut.kroyeri <- seewave::cutw(kroyeri, f=44100, from=0.2, to=1.1, output = "Wave")
```

```r
# Export ".wav" files containing acoustic units and store on previosly created folder
writeWave(cut.cuvieri, filename = file.path(wav.at, "cut.cuvieri.wav"), extensible = FALSE)
writeWave(cut.centralis, filename = file.path(wav.at, "cut.centralis.wav"), extensible = FALSE)
writeWave(cut.kroyeri, filename = file.path(wav.at, "cut.kroyeri.wav"), extensible = FALSE)
```


## 4. Define dimensions for the sound window

In order to secure a meaningful comparison of sound waves through semilandmark acquisition, eigensound analysis requires some standardization to `".wav"` files that would otherwise lead to errors or biased results (MacLeod et al., 2013; Rocha & Romano in prep). 

First, define the sound window dimensions that encompass the whole sample of acoustic units. These dimensions are represented by the time (*x*-axis) and frequency (*y*-axis) limits for spectrogram images, which are respectively defined by the `tlim` and `flim` arguments in `eigensound` function. 

Time limits should be based on the acoustic unit with longest duration within the sample, whereas frequency limits should consider the unit with largest frequency bandwidth. In the present sample study (Fig. 6), the longest units are also the ones with broader frequency bandwidths (*i.e.* `kroyeri` sample), with aproximately 0.7 s duration and highest frequencies close to 3.5 kHz. Therefore, the sound window dimensions that encompass the whole sample can be defined with `tlim = c(0, 0.8)` and `flim = c(0, 4)`.

This can be exemplified using `spectro` function from `seewave` package:

```{r Fig6-window-dimensions, fig.align='center', fig.height=5, fig.width=7.2}
# Spectrogram plots using standardized sound window dimensions
par(mfrow=c(2,2), mar=c(4,4,2,2))
seewave::spectro(cut.centralis, flim=c(0, 4), tlim=c(0, 0.8), main="data(centralis)",
                 wl=512, f=44100, ovlp=70, grid=FALSE, scale=FALSE)
seewave::spectro(cut.cuvieri, flim=c(0, 4), tlim=c(0, 0.8), main="data(cuvieri)", 
                 wl=512, f=44100, ovlp=70, grid=FALSE, scale=FALSE)
seewave::spectro(cut.kroyeri, flim=c(0, 4), tlim=c(0, 0.8), main="data(kroyeri)", 
                 wl=512, f=44100, ovlp=70, grid=FALSE, scale=FALSE)
```

**Figure 6:** Spectrogram images with standardized sound window dimensions. 


## 5. Alignment of acoustic units at the beginning of a sound window

The eigensound protocol also require acoustic units to be placed at the beginning of a sound window before the analysis. This ensure that variation in each semilandmark is due to energy shifts within the call, not to changes in their relative position in the sound window (MacLeod et al., 2013).

Although this arbitrary alignment could be performed on numerous softwares of acoustic analysis outside `R` platform (*e.g.* [Audacity](https://www.audacityteam.org/), [Raven Pro](http://ravensoundsoftware.com/software/raven-pro/)), `align.wave` function (`SoundShape` package) provide an easy alternative to automatically align the units at the beginning of a sound window whilst also standardizing the durations of `".wav"` files (see section 4.1). This prevents errors when running `eigensound` function (Rocha & Romano *in prep*). 

In order to verify the alignment, run `eigensound` with `analysis.type = "twoDshape"` and `plot.exp = TRUE`, which will create 2D spectrogram images and store them on the folder specified by `store.at` (see section 2 for folder paths), a helpful option for the verification of appropriate alignment and sound window dimensions.

Below is the code employed for the alignment of sound units and verification of sound window dimensions:
  
```r
# Place sounds at the beginning of a sound window
align.wave(wav.at=wav.at, wav.to="Aligned", time.length = 0.8)

# Verify alignment using analysis.type = "twoDshape"
eigensound(analysis.type = "twoDshape", wav.at = file.path(wav.at, "Aligned"),
           store.at=store.at, plot.exp=TRUE, flim=c(0, 4), tlim=c(0, 0.8))

# Go to folder specified by store.at and check jpeg files created
```

If either the alignment, or the sound window dimensions, are not ideal (*e.g.* units far from the beginning of sound window; sounds overlapping the edges of sound window), run `align.wave` with different values of `time.length` and/or `time.perc`, then use `eigensound` to verify the updated spectrogram outputs (see Rocha & Romano *in prep* for details). 

The ideal window dimensions and the alignment of units are often achieved after a few attempts. If this is troublesome, consider revisiting the relative amplitude (`dBlevel`) as the background noise could be interfering with `align.wave` (see section 6).


## 6. Set relative amplitude background

Next is the definition of a relative amplitude value (`dBlevel`) to be used as background in the 3D spectrogram (MacLeod et al., 2013). This is an iterative process that can be implemented by `eigensound` with `analysis.type = "twoDshape"` and `plot.exp = TRUE`, and should lead to spectrogram images with minimum influence from background noise (see Rocha & Romano *in prep* for details).

In the present study sample, the curve of relative amplitude was set at -25 dB (Fig. 7), which is expressed as an absolute value for `dBlevel` arguments in `SoundShape` functions (*i.e.* `dBlevel = 25`). 

The code below illustrate the `dBlevel` using `threeDspectro` function:
  
```{r Fig7-dBlevel-curve, fig.align='center', fig.height=3, fig.width=7.2}
# 2D spectrogram with curves of relative amplitude at -25 dB
par(mfrow=c(1,2), mar=c(4,4,1,1))
s.kro <- seewave::spectro(cut.kroyeri, flim=c(0, 4), tlim = c(0, 0.8),  
                          grid=F, scale=F, f=44100, wl=512, ovlp=70, cont=TRUE, 
                          contlevels = seq(-25, -25, 1), collevels = seq(-40, 0, 0.1))

# 3D spectrogram (with a lower dBlevel for illustrative purpuses)
threeDspectro(cut.kroyeri, dBlevel=40, flim=c(0, 4), tlim=c(0, 0.8), main="",
              colkey=list(plot=FALSE), cex.axis=0.4, cex.lab=0.8, resfac=2)

# Set background at -40 dB and remove -Inf values from spectrogram data 
for(i in 1:length(s.kro$amp)){if(s.kro$amp[i] == -Inf |s.kro$amp[i] <= -40)
{s.kro$amp[i] <- -40}}

# Add curve of relative amplitude
plot3D::contour3D(x=s.kro$time, y=s.kro$freq, colvar=t(s.kro$amp), z=-25,
                  plot=T, add=T, addbox=F, col="black", lwd=1.9, nlevels=2, dDepth=0.25)
```

**Figure 7:** 2D and 3D spectrograms (left and right, respectively) with relative amplitude contours highlighted by black lines (`dBlevel = 25`). Spectrogram images from `kroyeri` sample.


## 7. Define sampling grid and run `eigensound`

Once the sound window dimensions are defined (section 4), the acoustic units placed at the beginning of a sound window (section 5), and the relative amplitude background is set (section 6), next is the definition of sampling grid dimensions that will be used for semilandmark acquisition (*i.e.* number of cells per side; `x.length` and `y.length` arguments, `eigensound` function; Fig. 8). 

In our study sample, we opted for 70 cells on the time (*x*-axis, `x.length = 70`) and 47 cells on the frequency (*y*-axis, `y.length = 47`), which was iteratively defined with the aid of `threeDspectro` function, as exemplified below:
  
```{r Fig8-sampling-grid, fig.align='center', fig.height=4, fig.width=7.2}
# Using threeDspectro to visualize sampling grid 
par(mfrow=c(1,2), mar=c(1,2,1,0)) 

# As "surface"
threeDspectro(cut.kroyeri, samp.grid=TRUE, x.length=70, y.length=47, plot.type="surface", 
              dBlevel=25, flim=c(0, 4), tlim=c(0, 0.8), f=44100, wl=512, ovlp=70, main="As 'surface'",
              colkey=list(plot=FALSE), cex.axis=0.4, cex.lab=0.8)

# As "points"
threeDspectro(cut.kroyeri, samp.grid=TRUE, x.length=70, y.length=47, plot.type="points", 
              dBlevel=25, flim=c(0, 4), tlim=c(0, 0.8), f=44100, wl=512, ovlp=70, main="As 'points'",
              colkey=list(plot=FALSE), cex.axis=0.4, cex.lab=0.8)

```

**Figure 8:** Spectrogram data as (left) simplified surface, and (right) colored semilandmarks acquired from the intersections of sampling grid (*i.e.* `x.length=70` and ` y.length=47`). Spectrogram images from `kroyeri` sample.


### 7.1 Run `eigensound` function

It is now possible to acquire comparable semilandmark coordinates using `eigensound` function. Results can be simultaneosly assigned to an `R` object, and/or stored as the native file format of TPS series ([Rohlf, 2015](http://www.italian-journal-of-mammalogy.it/The-tps-series-of-software,77186,0,2.html)), a `".tps"` file to be used by numerous softwares of geometric analysis of shape. Herein, we focus on the analysis within `R` platform, so the results are assigned to the `R` object `eig.sample`, which is available as sample data from `SoundShape`.

**Note:** `eig.sample` comprises all vocalizations present in the samples of `centralis`, `cuvieri` and `kroyeri`, which led to three acoustic units per species; a total of nine `".wav"` files stored in the same folder. Use `help(eig.sample)` or check Rocha & Romano (*in prep*) for details. 

In the following code, `eigensound` is run with a logarithmic scale on the time axis (*i.e.* `log.scale = TRUE`; see Rocha & Romano *in prep* for details):
  
```r
# Sample semilandmarks for each ".wav" file on a folder using a logarithmic sampling grid
# Export 3D graphs with semilandmarks as colored points for inspection
eig.sample <- eigensound(analysis.type="threeDshape", dBlevel=25, 
                         f=44100, wl=512, ovlp=70, flim=c(0, 4), tlim=c(0, 0.8),
                         x.length=70, y.length=47, log.scale=TRUE, plot.exp=TRUE, plot.type="points",
                         wav.at=file.path(wav.at, "Aligned"), store.at=store.at)

# Go to folder specified by store.at and check jpeg files created
```

## 8. Principal Components Analysis

After employing a sampling grid to acquire semilandmarks from sound waves (section 7), the eigensound protocol proceeds to a dimensionality reduction procedure that facilitate comparison of sound shape data. Herein, we opted for a Principal Components Analysis (PCA), which allow complex sound waves to be described and plotted onto major axes (PCs) encompassing the majority of variance within the sample (MacLeod et al., 2013).

The PCA can be performed using `prcomp` function (`stats` package), as exemplified below:
  
```{r pca}
# PCA using three-dimensional semilandmark coordinates embeeded in eig.sample
pca.eig.sample <- stats::prcomp(geomorph::two.d.array(eig.sample))

# View summary results
summary(pca.eig.sample)
```

**Note:** At this point, consider employing a stopping rule to select which PCs should be retained as nontrivial and interpretable, and which ones should be ignored (*e.g.* broken stick models, [vegan](https://CRAN.R-project.org/package=vegan) package) (Jackson, 1993; see Rocha & Romano *in prep* for details). 


### 8.1 Hypothetical sound shape configurations from semilandmark data

Before proceeding to the ordination of Principal Components (PCs), the eigensound protocol also includes hypothetical sound surfaces to be interpreted along with the ordination plots (MacLeod et al., 2013). These surfaces are calculated from the sample and represent the variation embedded in each PC axis, therefore enhancing the visualization and comprehension on how sound shape changed along each of the PC scores.

In `SoundShape` package, the hypothetical sound shapes can be created using `hypo.surf` function, which enables the calculation of either the mean shape configuration from the sample (*i.e.* consensus shape; Zelditch et al., 2012), or minimum and maximum deformations relative to PCs, as exemplified below:
  
```{r Fig9-mean-surface, fig.align='center', fig.height=3, fig.width=7.2}
# Create hypothetical sound surfaces using hypo.surf

# Mean shape configuration (consensus)
hypo.surf(eig.sample,  PC="mean", flim=c(0, 4), tlim=c(0, 0.8), x.length=70, y.length=47,
          cex.lab=0.7, cex.axis=0.5, cex.main=1)
```

**Figure 9:** Hypothetical sound surface (acquired using `hypo.surf` function) representing mean shape configuration from `eig.sample` sample of data.

```{r Fig10-PC1-surface, fig.align='center', fig.height=3, fig.width=7.2}
# Minimum and maximum deformations - Principal Component 1
hypo.surf(eig.sample, PC=1, flim=c(0, 4), tlim=c(0, 0.8), x.length=70, y.length=47,
          cex.lab=0.7, cex.axis=0.5, cex.main=1)
```

**Figure 10:** Hypothetical sound surfaces (acquired using `hypo.surf` function) representing minimum and maximum deformations relative to PC1 in the PCA featuring `eig.sample` sample of data.

```{r Fig11-PC2-surface, fig.align='center', fig.height=3, fig.width=7.2}
# Minimum and maximum deformations - Principal Component 2
hypo.surf(eig.sample, PC=2, flim=c(0, 4), tlim=c(0, 0.8), x.length=70, y.length=47,
          cex.lab=0.7, cex.axis=0.5, cex.main=1)
```

**Figure 11:** Hypothetical sound surfaces (acquired using `hypo.surf` function) representing minimum and maximum deformations relative to PC2 in the PCA featuring `eig.sample` sample of data.


### 8.2 Ordination plot with `pca.plot` function

Among the benefits of employing a PCA on multivariate data is the possibility to generate ordination plots encompassing the majority of variation embeeded in the sample (Fig. 12). These plots simplify description and are widely employed in exploratory data analysis, specially when one is looking for potential groups within the sample  (Zelditch et al. 2012).

The ordination plot is facilitated by `pca.plot` function (`SoundShape` package), which require the output of a PCA performed by `prcomp` function (`stats` package) and a vector with `groups` to be colored. 

The code chunk below exemplifies how create an ordination plot using `pca.plot`:
  
```{r Fig12-ordination-plot, fig.align='center', fig.height=5, fig.width=5}
# PCA using semilandmark coordinates
pca.eig.sample <- stats::prcomp(geomorph::two.d.array(eig.sample))

# Verify names of acoustic units from sample 
dimnames(eig.sample)[[3]]

# Based on those names, create factor to use as groups in subsequent ordination plot
sample.gr <- factor(c(rep("centralis", 3), rep("cuvieri", 3), rep("kroyeri", 3)))

# Ordination plot
pca.plot(pca.eig.sample, groups=sample.gr, conv.hulls=sample.gr, leg.pos="bottomright", cex=1.2)
```

**Figure 12:** Ordination plot using `eig.sample` data acquired from the samples of `centralis`, `cuvieri` and `kroyeri`. 


## 9. Interpreting the outputs of `SoundShape`

In order to fully comprehend how sound shape changes along the studied sample, the PCA outcome should be interpreted along with the visualization of hypothetical sound shapes (Figs. 9 -- 11) and the ordination plot (Fig. 12).

The ordination plot (Fig. 12) represent 89.7% of the whole variance in our dataset, which yielded a clear structuring of units from different species. In addition, the hypothetical sound surfaces from the main axis of variation (*i.e.* mean shape and PCs) clearly represented the sound shapes of acoustic units employed in the study.

The higher positive values of PC1, for instance, corresponded to acoustic units with clear harmonic structure and broad frequency bandwidth (Fig. 10), a hypothetical sound shape remarkably similar to three-dimensional spectrograms from `kroyeri` sample. Not coincidently, the units from `kroyeri` scored high positive PC1 values. Lower and negative values of PC1, on the other hand, were less obvious, with a hypothetical shape that gather sonic information from `cuvieri` and `centralis` samples, both species with negative PC1 scores. A similar pattern is observed in PC2 axis (Fig. 11), with positive PC2 values referring to broad frequency bandwidth and no harmonic structure  (*i.e.* `centralis` sample), and negative PC2 values representing short durations and clear harmonic structure  (*i.e.* `cuvieri` sample). 


## References
Jackson, D. A. (1993). Stopping rules in Principal Components Analysis: A comparison of heuristical and statistical approaches. *Ecology, 74*(8), 2204-2214.

MacLeod, N., Krieger, J. & Jones, K. E. (2013). Geometric morphometric approaches to acoustic signal analysis in mammalian biology. *Hystrix, the Italian Journal of Mammalogy, 24*(1), 110-125. doi: 10.4404/hystrix-24.1-6299

Rocha, P. & Romano, P. (*in prep*) The shape of sound: A new `R` package that crosses the bridge between Bioacoustics and Geometric Morphometrics.

Rohlf, F.J. (2015) The tps series of software. *Hystrix 26*, 9-12.

Zelditch, M. L., Swiderski, D. L., Sheets, H. D., & Fink, W. L. (2012). *Geometric morphometrics for biologists: A primer*. Elsevier (Second Edition). Elsevier, San Diego. 
